State-space analysis of a continuous gravitational wave source with a pulsar timing array: inclusion of the pulsar terms

### 2024-01-09 manuscript/manuscript_2/paper.tex
complete draft

% Abstract of the paper
\begin{abstract}	
	 Pulsar timing arrays can detect continuous nanohertz gravitational waves emitted by individual supermassive black hole binaries. The data analysis procedure can be formulated within a state-space framework, wherein intrinsic achromatic spin wandering is tracked using a Kalman filter concurrently with the pulse frequency modulation induced by a gravitational wave from a single source. The modulation is the sum of terms proportional to the gravitational wave strain at the Earth and at every pulsar in the array. Here we generalize previous state-space formulations of the pulsar timing array problem to include the pulsar terms. The performance of the generalized Kalman filter is tested using astrophysically representative software injections in Gaussian measurement noise. It is shown that including the pulsar terms corrects for biases in the parameter estimates (especially the sky position of the source) and decreases the root-mean-square relative error in the estimated one-dimensional posteriors by to 99\%, for a loud gravitational wave source with characteristic wave strain $h_0 = 10^{-12}$. For a quiet gravitational wave signal with $h_0 = 5 \times 10^{-15}$ there is no significant improvement in parameter accuracy, but the minimum detectable strain drops by $13\%$.
\end{abstract}

@@

\section{Validation with synthetic data}\label{sec:rep_example1}
In the rest of this paper, we test the performance of the PTA analysis scheme in K24, once it is generalized to include the pulsar terms in the inference model as described in Section 3. We refer the reader to the appendices for detailed instructions about how to implement the Kalman filter (Appendix \ref{sec:kalman}) and nested sampler (Appendix \ref{sec:nested_sampling}) and integrate them in a unified workflow (Appendix \ref{sec:workflow}). In this section, by way of preparation, we explain how to generate the synthetic data employed in the tests, namely noisy frequency time series $f_{\rm m}^{(n)}(t)$ for $1 \leq n \leq N$. Tests are performed for multiple random realizations of $\xi^{(n)}(t)$ in order to quantify the irreducible ``cosmic'' variance in the inference output (e.g.\ estimates of ${\boldsymbol{\theta}}_{\rm gw}$). Every real PTA analysis witnesses a unique realization of $\xi^{(n)}(t)$ --- the actual, astronomical one --- but there is no way to determine where this realization lies within the admissible statistical ensemble. \newline 
%[end of first para of sec 4]

@@

\begin{table*}
	\centering
	%	\resizebox{\textwidth}{!}{%
		%	\renewcommand{\arraystretch}{1.0} % Default value: 1
		\begin{tabular}{lccll}
			\toprule
			Set&Parameter & Injected value & Units & Prior  \\
			\hline
			\multirow{7}{2mm}{$\boldsymbol{\theta}_{\rm gw}$} & $\Omega$       & $5 \times 10^{-7}$ & Hz & LogUniform($10^{-9}$, $10^{-5}$) \\
			& $\alpha$          & $1.0$  & rad & Uniform($0, 2 \pi $)\\
			& $\delta$              & $1.0$  & rad & Uniform($-\pi/2, \pi/2$) \\
			& $\psi$              & $2.50$ & rad & Uniform($0, 2 \pi $) \\
			& $\Phi_0$          & $0.20$ & rad & Uniform($0, 2 \pi $) \\
			& $h_0$            & $5 \times 10^{-15}$ & --- & LogUniform($10^{-15}$, $10^{-9}$) \\
			& $\iota$             & $1.0$ & rad & Uniform($0, \pi$) \\ 
			\hline
			\vspace{1mm}& $f_{\rm em}^{(n)} (t_1)$       & $f_{\rm ATNF}^{(n)}$ & Hz & Uniform$\left[f_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{f}, f_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{f} \right]$ \\
      \multirow{2}{2mm}{$\boldsymbol{\theta'}_{\rm psr}$} & $\dot{f}_{\rm em}^{(n)} (t_1)$       & $\dot{f}_{\rm ATNF}^{(n)}$ & s$^{-2}$ & Uniform$\left[ \dot{f}_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{\dot{f}}, \dot{f}_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{\dot{f}} \right]$ \\
		     & $\sigma^{(n)}$              & $\sigma_{\rm SC}^{(n)}$ & $s^{-3/2}$ & LogUniform$ \left [10^{-2} \sigma_{\rm SC}^{(n)}, 10^2 \sigma_{\rm SC}^{(n)} \right ]$ \\
			& $\gamma^{(n)}$              & $10^{-13}$ & s$^{-1}$ & --- \\
			\vspace{1mm} &  $\chi^{(n)}$       &$\Omega \left[ 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}_{\rm ATNF} \right]  d_{\rm ATNF}^{(n)} $  & rad & Uniform($0, 2 \pi $) \\
			\bottomrule
		\end{tabular}
		\caption{Injected static parameters used to generate synthetic data to validate the analysis scheme including the pulsar-terms. The prior used for Bayesian inference is also displayed (rightmost column).  The top and bottom sections of the table contain $\boldsymbol{\theta}_{\rm gw}$ and $\boldsymbol{\theta'}_{\rm psr}$ respectively. The subscript ``ATNF'' denotes values obtained from the ATNF pulsar catalogue as described in Section \ref{sec:rep_example1}. The subscript ``SC'' on $\sigma^{(n)}$ indicates that the injected value is calculated from Equation \eqref{eq:sigmap_f} and the empirical timing noise model for MSPs in \protect \cite{Shannon2010}. The quantities $\eta^{(n)}_{f}$ and $\eta^{(n)}_{\dot{f}}$ are the uncertainties in $f^{(n)}_{\rm em} (t_1)$ and $\dot{f}^{(n)}_{\rm em} (t_1)$ respectively, as quoted in the ATNF catalogue. We do not infer infer $\gamma^{(n)} \sim 10^{-5} T_{\rm obs}$, so no prior is set. The priors on $\boldsymbol{\theta}_{\rm psr}$ are justified in Appendix \ref{sec:set_priors}.
		}
		\label{tab:parameters_and_priors}
	\end{table*}
In order to synthesize $\boldsymbol{Y}(t) = \{f_{\rm m}^{(1)}(t), \dots, f_{\rm m}^{(N)}(t) \}$, we integrate Equations \eqref{eq:frequency_evolution}--\eqref{eq:spinevol} numerically using a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}. This produces random realizations of $f_{\rm p}^{(n)}(t)$ for $1\leq n \leq N$, which we convert to $f_{\rm m}^{(n)}(t)$ via Equations \eqref{eq:measurement}--\eqref{eq:g_func_trig}. The numerical solutions depend on how we choose $\boldsymbol{\theta}_{\rm psr}$, ${\boldsymbol{q}}^{(n)}$ and $\sigma_{\rm m}$ or, equivalently, how we specify the configuration of a synthetic PTA. In Section  \ref{sec:pe_and_ms} we describe how we choose the remaining elements of $\boldsymbol{\theta}'$, namely $\boldsymbol{\theta}_{\rm gw}$. This latter step is equivalent to specifying the synthetic SMBHB source and differs from one test to the next according to the goal of the test.  \newline

@@

\section{Parameter estimation} \label{sec:pe_and_ms} 
In this section, we apply the Kalman filter and nested sampler to calculate the joint posterior probability distribution $p({\boldsymbol{\theta'}} | {\boldsymbol{Y}})$ and compare it to the known, injected values. The procedure is undertaken for multiple realisations of $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$, and we estimate $\boldsymbol{\theta'}$ independently for each realisation. The aims are (i) to demonstrate that the analysis scheme works, i.e.\ that it converges to a well-behaved, unimodal posterior for multiple noise realisations; (ii) to give a preliminary sense of its accuracy, and (iii) to quantify the natural random dispersion in the one-dimensional posterior medians. Quantifying the dispersion is important since it is a practical measure of the scheme's accuracy when it is applied to real astronomical data, where the true parameter values and specific noise realisations are unknown. The injected $\boldsymbol{\theta}_{\rm gw}$ values are selected to be astrophysically representative, as per the top section of Table \ref{tab:parameters_and_priors}.The static pulsar parameters $\boldsymbol{\theta'}_{\rm psr}$ are as specified in Section \ref{sec:rep_example1}. All injected static parameters $\boldsymbol{\theta}'$ are summarised in the second column of Table \ref{tab:parameters_and_priors}. The specification of the priors on $\boldsymbol{\theta'}$ is described in Appendix \ref{sec:set_priors} and summarised in the rightmost column of Table \ref{tab:parameters_and_priors}. \newline

@@

Figure \ref{fig:corner_plot_1} displays the posterior distribution of $\boldsymbol{\theta}_{\rm gw}$ for ten noise realisations in the form of a traditional corner plot for the representative SMBHB source parametrized as per the top portion of Table \ref{tab:parameters_and_priors}. The histograms are one-dimensional posteriors, marginalized over the six other parameters. Each coloured curve corresponds to a different noise realisation. The solid orange line marks the known injected value. The two-dimensional contours mark the (0.5, 1, 1.5, 2)-sigma level surfaces. All histograms and contours are consistent with a unimodal joint posterior, which peaks near the known, injected values. There is scant evidence of railing against the prior bounds. There is no strong evidence for correlations between parameter pairs, e.g.\ banana-shaped contours, although weak-correlations are evident between $\Omega$ and $\Phi_0$, $\psi$ and $\alpha$, and $\iota$ and $h_0$. The injected value is contained within the 90\% credible interval of the one-dimensional marginalized posteriors in 60 out of the 70 possible combinations of the seven parameters and 10 realizations. The injected value falls within the 90\% credible interval for $\Omega, \Phi_0, \psi$, and  $\iota$ 40 out of 40 times. There is appreciable natural dispersion in the one-dimensional posterior medians between realisations. The minimum and maximum coefficients of variation are 0.2\% and 52\% for $\Omega$ and $\Phi_0$ respectively. The remaining parameters have typical coefficients $\lessapprox 10 \%$. For this representative example, with characteristic strain $h_0 = 5 \times 10^{-15}$ the analysis scheme estimates $\boldsymbol{\theta}_{\rm gw}$ accurately. Analogous results for synthetic data with higher strain ($h_0 = 10^{-12}$) are presented in Appendix \ref{sec:app_high_SNR} (see Figure \ref{fig:corner_high_snr_appendix}). The dispersion between noise realisations decreases with increasing $h_0$ is reduced and the one-dimensional posteriors for $\iota$ and $h_0$ are more symmetric about the injected value. \newline

@@

The PP results are displayed in Figure \ref{fig:parameter_space}. The shaded grey contours enclose the $1\sigma$, $2\sigma$, and $3\sigma$ significance levels for 200 injections. The curves for the five parameters, colour-coded in the legend, are approximately linear. Some parameters are consistently well estimated across the parameter domain, e.g. the blue curve for $\Omega$ lies everywhere within the 2$\sigma$ shaded region. Conversely some parameters, e.g. $\Phi_0$, stray outside the $1\sigma$ shaded region and show evidence of being slightly over-constrained; there are more injections contained within lower-value credible intervals than would be expected statistically, and fewer injections contained within higher-value credible intervals. This occurs due to the low SNR of the injected GW signal and the varying sensitivity of the specific PTA configuration used to different regions of the sky. More quantitatively, for the best estimated parameter $\Omega$ the injection is contained within the 90 \% credible interval in 91\% of cases. For the worst estimated parameters such as $\Phi_0$, the injection is contained within the 90 \% credible interval in 81\% of cases. \newline

Manually inspecting the individual corner plots for each of the 200 injections, we see that the nested sampler does not return a unimodal posterior in the rare event that the source is located unfavourably, with $\boldsymbol{n} \cdot \boldsymbol{q}^{(n)} \approx -1$, such that one cannot infer $\chi^{(n)}$ accurately (see Section \ref{sec:chi_estim}). All injected sources are ``observed" with the same settings such as $T_{\rm obs}$, $T_{\rm cad}$ and $n_{\rm live}$. In Figure \ref{fig:parameter_space2} we display a second PP plot, arranged identically to Figure \ref{fig:parameter_space}, but assuming that the injected values of $\chi^{(n)}$ are known exactly for the sake of testing, i.e. setting a delta-function prior on $\chi^{(n)}$. The excursions out of the shaded region diminish compared to Figure \ref{fig:parameter_space}, confirming the importance of accurate estimates of $\chi^{(n)}$. For the best estimated parameter $\Omega$ the injection is contained within the 90 \% credible interval in 91\% of cases, the same as in Figure \ref{fig:parameter_space}. For the worst estimated parameters such as $\Phi_0$, the injection is contained within the 90 \% credible interval in 89\% of cases, an improvement over the results of Figure Figure \ref{fig:parameter_space}. \newline

@@

\section{Bias when neglecting the pulsar terms}\label{sec:earth_vs_psr}
In Section \ref{sec:rep_example1} we have validated the state-space analysis scheme when the pulsar terms are included in the inference model. In this section, critically, we compare the effect of including the pulsar-term in the model (i.e. using Equation \eqref{eq:g_func_trig_chi}), as opposed to the omitting the pulsar-term (i.e. Equation \eqref{eq:g_func_trig_earth}). We do this using the specific representative case of the individual quasi-monochromatic SMBHB source described in Table \ref{tab:parameters_and_priors}. We refer to the model where the pulsar-term is included alongside the Earth term as $\mathcal{M}_{\rm psr \& earth}$. We refer to the model where the pulsar-term is omitted as 
$\mathcal{M}_{\rm earth}$. Model $\mathcal{M}_{\rm psr \& earth}$ is parameterised by $\boldsymbol{\theta}'$. Model $\mathcal{M}_{\rm earth}$ is parameterised by $\boldsymbol{\theta}_{\rm earth} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta}'_{\rm psr}$. Both models are applied to identical realisations of the data, generated using the procedure as described in Section  \ref{sec:rep_example1}. The priors on the static parameters in each model are also identical, with the addition of a uniform prior on $\chi^{(n)}$ for $\mathcal{M}_{\rm psr \& earth}$, as described in Appendix \ref{sec:set_priors}. \newline

In Section \ref{sec:psr_v_earth_pe} we compare the accuracy of the estimates of $\boldsymbol{\theta'}$, $\boldsymbol{\theta}_{\rm earth}$ returned by the respective models. In Section \ref{sec:psr_v_earth_bayes} we compare the minimal detectable GW strain for the two models by calculating the model evidence and comparing it to the evidence for a null model that does not contain a GW.

@@

The results for the seven parameters of  $\boldsymbol{\theta}_{\rm gw}$ are shown in Figure \ref{fig:corner_plot_compare_high} for the high-SNR system and in Figure \ref{fig:corner_plot_compare_low} for the low-SNR system. The corner plot is arranged identically to Figure \ref{fig:corner_plot_1}, however the different coloured curved now correspond to different inference models, rather than different realisations of $\boldsymbol{Y}$. The blue curves are the results derived using $\mathcal{M}_{\rm earth}$. The green curves are the results derived using $\mathcal{M}_{\rm psr \& earth}$. \newline

For the high-SNR case, Figure \ref{fig:corner_plot_compare_high}, the one-dimensional posteriors for $\mathcal{M}_{\rm earth}$ are evidently biased, as was observed in K24, as well as \cite{Zhupulsarterms} and \cite{Chen2022}. Particular biases are observed in $\psi, \iota $ and $\alpha$, with the blue posterior displaced with respect to the orange injection line. For example, for $\iota$, whilst the injected value is contained within the 90\% credible interval, the median of the posterior of the Earth-term model is shifted from the injected value by 0.35 radians. The inclusion of the pulsar terms corrects for this bias; the green one-dimensional posteriors exhibit no bias and are generally symmetric about the injected value. Moreover, for every pair of parameters the injection values are contained within the 2-sigma equivalent contours for the $\mathcal{M}_{\rm psr \& earth}$, which is not true for $\mathcal{M}_{\rm earth}$. \newline

>>>>

% Abstract of the paper
\begin{abstract}	
	 Pulsar timing arrays can detect continuous nanohertz gravitational waves emitted by individual supermassive black hole binaries. The data analysis procedure can be formulated within a state-space framework, wherein intrinsic achromatic spin wandering is tracked using a Kalman filter concurrently with the pulse frequency modulation induced by a gravitational wave from a single source. The modulation is the sum of terms proportional to the gravitational wave strain at the Earth and at every pulsar in the array. Here we generalize previous state-space formulations of the pulsar timing array problem to include the pulsar terms. The performance of the generalized Kalman filter is tested using astrophysically representative software injections in Gaussian measurement noise. It is shown that including the pulsar terms corrects for biases in the parameter estimates (especially the sky position of the source) and decreases the root-mean-square relative error in the estimated one-dimensional posteriors by to 99\%, for a loud gravitational wave source with characteristic wave strain $h_0 = 10^{-12}$. For a quiet gravitational wave signal with $h_0 = 5 \times 10^{-15}$ there is no significant improvement in parameter accuracy, but the minimum detectable strain decreases by $13\%$.
\end{abstract}

@@

\section{Validation with synthetic data}\label{sec:rep_example1}
In the rest of this paper, we test the performance of the PTA analysis scheme in K24, once it is generalized to include the pulsar terms in the inference model as described in Section \ref{sec:pulsar_term}. We refer the reader to the appendices for detailed instructions about how to implement the Kalman filter (Appendix \ref{sec:kalman}) and nested sampler (Appendix \ref{sec:nested_sampling}) and integrate them in a unified workflow (Appendix \ref{sec:workflow}). In this section, by way of preparation, we explain how to generate the synthetic data employed in the tests, namely noisy frequency time series $f_{\rm m}^{(n)}(t)$ for $1 \leq n \leq N$. Tests are performed for multiple random realizations of $\xi^{(n)}(t)$ in order to quantify the irreducible ``cosmic'' variance in the inference output (e.g.\ estimates of ${\boldsymbol{\theta}}_{\rm gw}$). Every real PTA analysis witnesses a unique realization of $\xi^{(n)}(t)$ --- the actual, astronomical one --- but there is no way to determine where this realization lies within the admissible statistical ensemble. \newline 
%[end of first para of sec 4]

@@

\begin{table*}
	\centering
	%	\resizebox{\textwidth}{!}{%
		%	\renewcommand{\arraystretch}{1.0} % Default value: 1
		\begin{tabular}{lccll}
			\toprule
			Set&Parameter & Injected value & Units & Prior  \\
			\hline
			\multirow{7}{2mm}{$\boldsymbol{\theta}_{\rm gw}$} & $\Omega$       & $5 \times 10^{-7}$ & Hz & LogUniform($10^{-9}$, $10^{-5}$) \\
			& $\alpha$          & $1.0$  & rad & Uniform($0, 2 \pi $)\\
			& $\delta$              & $1.0$  & rad & Uniform($-\pi/2, \pi/2$) \\
			& $\psi$              & $2.50$ & rad & Uniform($0, 2 \pi $) \\
			& $\Phi_0$          & $0.20$ & rad & Uniform($0, 2 \pi $) \\
			& $h_0$            & $5 \times 10^{-15}$ & --- & LogUniform($10^{-15}$, $10^{-9}$) \\
			& $\iota$             & $1.0$ & rad & Uniform($0, \pi$) \\ 
			\hline
			& $f_{\rm em}^{(n)} (t_1)$       & $f_{\rm ATNF}^{(n)}$ & Hz & Uniform$\left[f_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{f}, f_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{f} \right]$ \\
       & $\dot{f}_{\rm em}^{(n)} (t_1)$       & $\dot{f}_{\rm ATNF}^{(n)}$ & s$^{-2}$ & Uniform$\left[ \dot{f}_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{\dot{f}}, \dot{f}_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{\dot{f}} \right]$ \\
		     $\boldsymbol{\theta'}_{\rm psr}$ & $\sigma^{(n)}$              & $\sigma_{\rm SC}^{(n)}$ & $s^{-3/2}$ & LogUniform$ \left [10^{-2} \sigma_{\rm SC}^{(n)}, 10^2 \sigma_{\rm SC}^{(n)} \right ]$ \\
			& $\gamma^{(n)}$              & $10^{-13}$ & s$^{-1}$ & --- \\
			\vspace{1mm} &  $\chi^{(n)}$       &$\Omega \left[ 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}_{\rm ATNF} \right]  d_{\rm ATNF}^{(n)} $  & rad & Uniform($0, 2 \pi $) \\
			\bottomrule
		\end{tabular}
		\caption{Injected static parameters used to generate synthetic data to validate the analysis scheme including the pulsar-terms. The prior used for Bayesian inference is also displayed (rightmost column).  The top and bottom sections of the table contain $\boldsymbol{\theta}_{\rm gw}$ and $\boldsymbol{\theta'}_{\rm psr}$ respectively. The subscript ``ATNF'' denotes values obtained from the ATNF pulsar catalogue as described in Section \ref{sec:rep_example1}. The subscript ``SC'' on $\sigma^{(n)}$ indicates that the injected value is calculated from Equation \eqref{eq:sigmap_f} and the empirical timing noise model for MSPs in \protect \cite{Shannon2010}. The quantities $\eta^{(n)}_{f}$ and $\eta^{(n)}_{\dot{f}}$ are the uncertainties in $f^{(n)}_{\rm em} (t_1)$ and $\dot{f}^{(n)}_{\rm em} (t_1)$ respectively, as quoted in the ATNF catalogue. We do not infer infer $\gamma^{(n)} \sim 10^{-5} T_{\rm obs}$, so no prior is set. The priors on $\boldsymbol{\theta}_{\rm psr}$ are justified in Appendix \ref{sec:set_priors}.
		}
		\label{tab:parameters_and_priors}
	\end{table*}
In order to synthesize $\boldsymbol{Y}(t) = \{f_{\rm m}^{(1)}(t), \dots, f_{\rm m}^{(N)}(t) \}$, we integrate Equations \eqref{eq:frequency_evolution}--\eqref{eq:spinevol} numerically using a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}. This produces random realizations of $f_{\rm p}^{(n)}(t)$ for $1\leq n \leq N$, which we convert to $f_{\rm m}^{(n)}(t)$ via Equations \eqref{eq:measurement}--\eqref{eq:g_func_trig}. The numerical solutions depend on how we choose $\boldsymbol{\theta}_{\rm psr}$, ${\boldsymbol{q}}^{(n)}$ and $\sigma_{\rm m}$ or, equivalently, how we specify the configuration of a synthetic PTA. In Section  \ref{sec:pe_and_ms} we describe how we choose the remaining elements of $\boldsymbol{\theta}'$, namely $\boldsymbol{\theta}_{\rm gw}$. This latter step is equivalent to specifying the synthetic SMBHB source and differs from one test to the next according to the goal of the test.  \newline

@@

\section{Parameter estimation} \label{sec:pe_and_ms} 
In this section, we apply the Kalman filter and nested sampler to calculate the joint posterior probability distribution $p({\boldsymbol{\theta'}} | {\boldsymbol{Y}})$ and compare it to the known, injected values. The procedure is undertaken for multiple realisations of $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$, and we estimate $\boldsymbol{\theta'}$ independently for each realisation. The aims are (i) to demonstrate that the analysis scheme works, i.e.\ that it converges to a well-behaved, unimodal posterior for multiple noise realisations; (ii) to give a preliminary sense of its accuracy, and (iii) to quantify the natural random dispersion in the one-dimensional posterior medians. Quantifying the dispersion is important since it is a practical measure of the scheme's accuracy when it is applied to real astronomical data, where the true parameter values and specific noise realisations are unknown. The injected $\boldsymbol{\theta}_{\rm gw}$ values are selected to be astrophysically representative, as per the top section of Table \ref{tab:parameters_and_priors}. The static pulsar parameters $\boldsymbol{\theta'}_{\rm psr}$ are as specified in Section \ref{sec:rep_example1}. All injected static parameters $\boldsymbol{\theta}'$ are summarised in the second column of Table \ref{tab:parameters_and_priors}. The specification of the priors on $\boldsymbol{\theta'}$ is described in Appendix \ref{sec:set_priors} and summarised in the rightmost column of Table \ref{tab:parameters_and_priors}. \newline

@@

Figure \ref{fig:corner_plot_1} displays the posterior distribution of $\boldsymbol{\theta}_{\rm gw}$ for ten noise realisations in the form of a traditional corner plot for the representative SMBHB source parametrized as per the top portion of Table \ref{tab:parameters_and_priors}. The histograms are one-dimensional posteriors, marginalized over the six other parameters. Each coloured curve corresponds to a different noise realisation. The solid orange line marks the known injected value. The two-dimensional contours mark the (0.5, 1, 1.5, 2)-sigma level surfaces. All histograms and contours are consistent with a unimodal joint posterior, which peaks near the known, injected values. There is scant evidence of railing against the prior bounds. There is no strong evidence for correlations between parameter pairs, e.g.\ banana-shaped contours, although weak-correlations are evident between $\Omega$ and $\Phi_0$, $\psi$ and $\alpha$, and $\iota$ and $h_0$. The injected value is contained within the 90\% credible interval of the one-dimensional marginalized posteriors in 60 out of the 70 possible combinations of the seven parameters and 10 realizations. The injected value falls within the 90\% credible interval for $\Omega, \Phi_0, \psi$, and  $\iota$ 40 out of 40 times. There is appreciable natural dispersion in the one-dimensional posterior medians between realisations. The minimum and maximum coefficients of variation are 0.2\% and 52\% for $\Omega$ and $\Phi_0$ respectively. The remaining parameters have typical coefficients $\lessapprox 10 \%$. For this representative example, with characteristic strain $h_0 = 5 \times 10^{-15}$ the analysis scheme estimates $\boldsymbol{\theta}_{\rm gw}$ accurately. Analogous results for synthetic data with higher strain ($h_0 = 10^{-12}$) are presented in Appendix \ref{sec:app_high_SNR} (see Figure \ref{fig:corner_high_snr_appendix}); the dispersion between noise realisations decreases with increasing $h_0$ and the one-dimensional posteriors for $\iota$ and $h_0$ are more symmetric about the injected value. \newline

@@

The PP results are displayed in Figure \ref{fig:parameter_space}. The shaded grey contours enclose the $1\sigma$, $2\sigma$, and $3\sigma$ significance levels for 200 injections. The curves for the five parameters, colour-coded in the legend, are approximately linear. Some parameters are consistently well estimated across the parameter domain, e.g. the blue curve for $\Omega$ lies everywhere within the 2$\sigma$ shaded region. Conversely some parameters, e.g. $\Phi_0$, stray outside the $1\sigma$ shaded region and show evidence of being slightly over-constrained; there are more injections contained within lower-value credible intervals than would be expected statistically, and fewer injections contained within higher-value credible intervals. This occurs due to the low SNR of the injected GW signal and the varying sensitivity of the specific PTA configuration used to different regions of the sky. More quantitatively, for the best estimated parameter $\Omega$ the injection is contained within the 90\% credible interval in 91\% of cases. For the worst estimated parameters such as $\Phi_0$, the injection is contained within the 90 \% credible interval in 81\% of cases. \newline

Manually inspecting the individual corner plots for each of the 200 injections, we see that the nested sampler does not return a unimodal posterior in the rare event that the source is located unfavourably, with $\boldsymbol{n} \cdot \boldsymbol{q}^{(n)} \approx -1$, such that one cannot infer $\chi^{(n)}$ accurately (see Section \ref{sec:chi_estim}). All injected sources are ``observed" with the same settings such as $T_{\rm obs}$, $T_{\rm cad}$ and $n_{\rm live}$. In Figure \ref{fig:parameter_space2} we display a second PP plot, arranged identically to Figure \ref{fig:parameter_space}, but assuming that the injected values of $\chi^{(n)}$ are known exactly for the sake of testing, i.e. setting a delta-function prior on $\chi^{(n)}$. The excursions out of the shaded region diminish compared to Figure \ref{fig:parameter_space}, confirming the importance of accurate estimates of $\chi^{(n)}$. For the best estimated parameter $\Omega$ the injection is contained within the 90\% credible interval in 91\% of cases, the same as in Figure \ref{fig:parameter_space}. For the worst estimated parameters such as $\Phi_0$, the injection is contained within the 90 \% credible interval in 89\% of cases, an improvement over the results of Figure \ref{fig:parameter_space}. \newline

@@

\section{Bias when neglecting the pulsar terms}\label{sec:earth_vs_psr}
In Section \ref{sec:pe_and_ms} we have validated the state-space analysis scheme when the pulsar terms are included in the inference model. In this section, critically, we compare the effect of including the pulsar-term in the model, i.e. using Equation \eqref{eq:g_func_trig_chi}, as opposed to the omitting the pulsar-term, i.e. using Equation \eqref{eq:g_func_trig_earth}. We do this using the specific representative case of the individual quasi-monochromatic SMBHB source described in Table \ref{tab:parameters_and_priors}. We refer to the model where the pulsar-term is included alongside the Earth term as $\mathcal{M}_{\rm psr \& earth}$. We refer to the model where the pulsar-term is omitted as 
$\mathcal{M}_{\rm earth}$. Model $\mathcal{M}_{\rm psr \& earth}$ is parameterised by $\boldsymbol{\theta}'$. Model $\mathcal{M}_{\rm earth}$ is parameterised by $\boldsymbol{\theta}_{\rm earth} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta}'_{\rm psr}$. Both models are applied to identical realisations of the data, generated using the procedure as described in Section  \ref{sec:rep_example1}. The priors on the static parameters in each model are also identical, with the addition of a uniform prior on $\chi^{(n)}$ for $\mathcal{M}_{\rm psr \& earth}$, as described in Appendix \ref{sec:set_priors}. \newline

In Section \ref{sec:psr_v_earth_pe} we compare the accuracy of the estimates of $\boldsymbol{\theta'}$ and $\boldsymbol{\theta}_{\rm earth}$ returned by the respective models. In Section \ref{sec:psr_v_earth_bayes} we compare the minimal detectable GW strain for the two models by calculating the model evidence and comparing it to the evidence for a null model that does not contain a GW.

@@

The results for the seven parameters of  $\boldsymbol{\theta}_{\rm gw}$ are shown in Figure \ref{fig:corner_plot_compare_high} for the high-SNR system and in Figure \ref{fig:corner_plot_compare_low} for the low-SNR system. The corner plot is arranged identically to Figure \ref{fig:corner_plot_1}, however the different coloured curves and contours now correspond to different inference models, rather than different realisations of $\boldsymbol{Y}$. The blue curves are the results derived using $\mathcal{M}_{\rm earth}$. The green curves are the results derived using $\mathcal{M}_{\rm psr \& earth}$. \newline

For the high-SNR case, Figure \ref{fig:corner_plot_compare_high}, the one-dimensional posteriors for $\mathcal{M}_{\rm earth}$ are evidently biased, as was observed in K24, as well as \cite{Zhupulsarterms} and \cite{Chen2022}. Particular biases are observed in $\psi, \iota $ and $\alpha$, with the blue posterior displaced with respect to the orange injection line. For example, considering $\iota$, whilst the injected value is contained within the 90\% credible interval, the median of the posterior of the Earth-term model is shifted from the injected value by 0.35 radians. The inclusion of the pulsar terms corrects for this bias; the green one-dimensional posteriors exhibit no bias and are generally symmetric about the injected value. Moreover, for every pair of parameters the injection values are contained within the 2-sigma equivalent contours for the $\mathcal{M}_{\rm psr \& earth}$, which is not true for $\mathcal{M}_{\rm earth}$. \newline

We define a root-mean-square relative error to quantify the accuracy of the one-dimensional posteriors with respect to the injection value as
\begin{equation}
\Delta_{\rm M}(\theta) = \sqrt{\frac{1}{K}\sum_{k=1}^{K} \frac{ \left[p_k(\theta) - \hat{\theta}\right]^2}{\hat{\theta}}} \ , \label{eq:mean_rel_error}
\end{equation}
where $p_k(\theta)$ is the $k$-th one-dimensional posterior sample for parameter $\theta$, $K$ is the total number of samples, $\hat{\theta}$ is the true injection value, and the subscript ${\rm M} \in \{ \text{earth, psr \& earth} \}$ indicates whether the posterior is estimated using Equation \eqref{eq:g_func_trig_earth} or Equation \eqref{eq:g_func_trig} respectively. The dependence of $p_k(\theta)$ on ${\rm M}$ is implied. The values of $\Delta_{\rm M}(\theta)$ for each parameter in Figure \ref{fig:corner_plot_compare_high} are summarised in Table \ref{tab:posterior_errors}. The difference in the error between the two models, $\Delta_{\rm earth}(\theta)-\Delta_{\rm psr \& earth}(\theta)$ is also presented. For every parameter, $\Delta_{\rm earth}(\theta)-\Delta_{\rm psr \& earth}(\theta)$ is positive, indicating that the parameter estimates are more accurate when including the pulsar terms, i.e. $\Delta_{\rm earth}(\theta) > \Delta_{\rm psr \& earth}$. In some cases the difference is modest;  $\Omega$ is recovered with high accuracy by both models such that $\Omega$, $\Delta_{\rm earth}(\Omega)-\Delta_{\rm psr \& earth}(\Omega) = 6.7 \times 10^{-3}$ rad. The improvement from using the pulsar terms is largest for $\iota$, with $\Delta_{\rm earth}(\iota) - \Delta_{\rm psr \& earth}(\iota)  = 0.24$ rad. As a percentage change, i.e. $1 - \Delta_{\rm earth}(\theta)/\Delta_{\rm psr \& earth}(\theta)$, the improvement from using $\mathcal{M}_{\rm psr \& earth}$ over $\mathcal{M}_{\rm earth}$ ranges from 61\% for $\Omega$ to 99\% for $\psi$ and $\alpha$. These improvements are consistent across noise realisations. \newline

For the low-SNR case,  Figure \ref{fig:corner_plot_compare_low}, there is less of a notable improvement in the parameter estimates. Qualitatively, the green and blue one-dimensional posteriors are generally overlaid. The two-dimensional contours suggest a slight improvement through using $\mathcal{M}_{\rm psr \& earth}$, but the difference is modest. The root-mean-square relative error $\Delta_{\rm M}(\theta)$ for the low SNR case is also reported in Table \ref{tab:posterior_errors}. We see that the error is larger for every element in $\boldsymbol{\theta}_{\rm gw}$ than in the high SNR case, as expected. The inclusion of the pulsar terms improve the estimates for five out of the seven static parameters; $\Delta_{\rm earth}(\theta)-\Delta_{\rm psr \& earth}(\theta)$ is positive for all parameters except $\psi$ and $h_0$. However it is hard to draw strong conclusions; for different realisations of the noise different parameters exhibit either improvement or deterioration in $\Delta_{\rm M}(\theta)$. We conclude that the inclusion of the pulsar terms does not significantly improve the parameter estimates in the low-SNR case. This is intuitive since the system is highly noise dominated such that an increasing model complexity does not aid parameter estimates. Whilst the inclusion of the pulsar terms only modestly improve the accuracy of the parameter estimates in the case of low-SNR, we will show in Section \ref{sec:psr_v_earth_bayes} that they are important for increasing the detection probability for a given signal strength. \newline
